---
title: "GIS Librarian Directory"
subtitle: "All verified GIS librarians and specialists at US R1 universities"
date: today
date-format: "MMMM D, YYYY"
---

```{r setup, include=FALSE}
library(tidyverse)
library(reactable)
library(htmltools)

people <- read_csv("../output/verified.csv", show_col_types = FALSE) %>%
  mutate(across(where(is.character), ~replace_na(., "")))
```

This table lists all **`r nrow(people)` GIS librarians and specialists** identified at Carnegie R1 research institutions. Use the search box and column filters to explore. Click a row to expand notes.

**Privacy note:** Entries include only information already public on institutional library staff directories (name, title, institution). Email addresses are not published here. To request a correction or removal from this directory, contact Tim Dennis at tdennis@library.ucla.edu or open an issue on [GitHub](https://github.com/jt14den/gis-librarians-library/issues).

```{r directory-table}
reactable(
  people %>%
    select(Name, `Title/Role`, Institution, Department = `Department/Unit`,
           `Profile URL`, `Last-Checked Date`, Notes),
  searchable = TRUE,
  filterable = TRUE,
  sortable   = TRUE,
  striped    = TRUE,
  highlight  = TRUE,
  defaultPageSize = 25,
  columns = list(
    Name = colDef(minWidth = 150, sticky = "left"),
    `Title/Role` = colDef(minWidth = 200),
    Institution  = colDef(minWidth = 220),
    Department   = colDef(minWidth = 180),
    `Profile URL` = colDef(
      name = "Profile",
      minWidth = 80,
      cell = function(value) {
        if (nchar(value) > 0) tags$a(href = value, target = "_blank", "Link") else ""
      }
    ),
    `Last-Checked Date` = colDef(name = "Checked", minWidth = 100),
    Notes = colDef(minWidth = 200)
  )
)
```

---

*Data collected March 2026. Corrections and additions welcome, please open an issue on [GitHub](https://github.com/jt14den/gis-librarians-library).*
